---
title: "Basic Probability and Statistics with R"
author: "Roberto Molinari"
date: "`r Sys.Date()`"
output: html_document
---

# Introductions

This document introduces fundamental statistical and probability operations in R. The examples cover:

1. Basic mathematical calculations
2. Probability distributions
3. Summary statistics
4. Factor variables
5. Exploring built-in datasets

These concepts form the foundation for statistical computing and data analysis in R.

---

# 1. Simple Calculations

R can be used as a scientific calculator. It supports arithmetic operations, mathematical functions, and constants.

```{r}
2 + 2
4 / 2
3 * 2^(-0.8)
sqrt(2)
pi
log(2)
log(9, base = 3)
exp(1.1)
cos(sqrt(0.9))
```

## Explanation

- `+`, `-`, `*`, and `/` perform basic arithmetic.
- `^` raises a number to a power.
- `sqrt()` computes square roots.
- `pi` returns the mathematical constant π.
- `log()` computes logarithms. By default, the natural logarithm is used.
- `exp()` computes the exponential function \(e^x\).
- `cos()` computes the cosine of an angle measured in radians.

### Storing Results in Variables

Results can be assigned to objects using the assignment operator `<-`.

```{r}
z <- sqrt(0.9)
cos(z)
```

Here:

- `sqrt(0.9)` is stored in the object `z`.
- The value of `z` is then used as the argument of the cosine function.

---

# 2. Probability Distributions

R provides many probability distributions through a consistent naming convention.

## Distribution Function Prefixes

For a distribution called `xxx`:

| Prefix | Meaning |
|----------|----------|
| `dxxx` | Density or probability mass function |
| `pxxx` | Cumulative probability |
| `qxxx` | Quantile function |
| `rxxx` | Random number generation |

For example, the normal distribution uses:

- `dnorm()`
- `pnorm()`
- `qnorm()`
- `rnorm()`

---

## Example: Normal Distribution

Suppose test scores follow a normal distribution:

\[
X \sim N(70, 15^2)
\]

We want to calculate:

\[
P(X > 90)
\]

```{r}
pnorm(
  q = 90,
  mean = 70,
  sd = 15,
  lower.tail = FALSE
)
```

### Explanation

- `q = 90` specifies the value of interest.
- `mean = 70` is the population mean.
- `sd = 15` is the population standard deviation.
- `lower.tail = FALSE` returns the probability above 90.

This computes:

\[
P(X > 90)
\]

The same function can be called using positional arguments:

```{r}
pnorm(q = 90)
pnorm(90, 70, 15)
```

The first call uses the standard normal distribution \(N(0,1)\), while the second uses \(N(70,15^2)\).

---

## Density Function

```{r}
dnorm(90, mean = 70, sd = 15)
```

This computes the height of the normal density curve at \(x=90\).

Note that a density value is **not** the same as a probability.

---

## Quantiles

```{r}
qnorm(0.95, mean = 70, sd = 15)
```

This finds the score corresponding to the 95th percentile.

In other words, 95% of observations are expected to fall below this value.

---

## Random Number Generation

```{r}
set.seed(985)

z <- rnorm(
  mean = 70,
  sd = 15,
  n = 1000
)
```

### Explanation

- `rnorm()` generates random observations from a normal distribution.
- `n = 1000` specifies the sample size.
- `set.seed()` ensures that results are reproducible.

The object `z` now contains 1000 simulated test scores.

---

# 3. Summary Statistics

Consider the sample:

```{r}
x <- c(12, 15, 18, 20, 22, 25, 30)
```

---

## Measures of Center and Spread

```{r}
mean(x)
median(x)
var(x)
sd(x)
```

### Explanation

- `mean()` computes the arithmetic average.
- `median()` computes the middle observation.
- `var()` computes sample variance.
- `sd()` computes sample standard deviation.

These statistics summarize the center and variability of the data.

---

## Statistics for Simulated Data

```{r}
mean(z)
median(z)
var(z)
sd(z)
```

Because `z` was generated from a normal distribution with mean 70 and standard deviation 15, the sample statistics should be close to those values.

---

## Quantiles

```{r}
quantile(z, probs = 0.95)
qnorm(0.95, mean = 70, sd = 15)
```

### Comparison

- `quantile()` calculates the empirical 95th percentile from the simulated sample.
- `qnorm()` calculates the theoretical 95th percentile from the population distribution.

The values should be similar but not necessarily identical due to sampling variability.

---

## Interquartile Range

```{r}
IQR(x)
```

The interquartile range (IQR) measures spread using:

\[
IQR = Q_3 - Q_1
\]

where:

- \(Q_1\) = first quartile
- \(Q_3\) = third quartile

The IQR is resistant to outliers.

---

## Minimum and Maximum

```{r}
min(x)
max(x)
range(x)
```

### Explanation

- `min()` returns the smallest observation.
- `max()` returns the largest observation.
- `range()` returns both values.

---

## Summary Function

```{r}
summary(x)
```

This provides:

- Minimum
- First Quartile
- Median
- Mean
- Third Quartile
- Maximum

It is a quick overview of a numeric variable.

---

# 4. Factor Variables

Factors are used to represent categorical variables.

```{r}
colors